Genetic alterations and expression patterns of genes coding for FGF, FGFR, and FGFBP proteins in urothelial cancers

### 2024-12-16 report/markdown/report.Rmd
Report and figure update
Figures and Report text for analyses with DepMap cell lines

* Five genetic and transcrional collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (n = `r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (n = `r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers (n = `r nrow(bcan$clinic)`) [@Damrauer2022]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BCAN cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA, IMvigor, and BCAN cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, IMvigor, and BCAN cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Validation of expression of the genes of interest was done with immunohistochemistry data retrieved from the [Human Protein Atlas](https://www.proteinatlas.org/). Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.

@@

* Mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ were predominantly classified as missense mutations caused by single nucleotide polymorphisms (__Figure \@ref(fig:fig-mutation-type)__ and __\@ref(fig:fig-variant-type)__, __Table \@ref(tab:tab-fgfr-class)__).  For _FGFR1_, the mutation hot-spots were located in the Ig-like domain 1 and kinase domain. The kinase domain was the prime target of mutations of _FGFR2_. For _FGFR3_, they affected protein residues were located between Ig-like domains 1 and 2 responsible for ligand biding, or at the membrane-spanning hinge between the ligand-binding region and the kinase domain responsible for activation of the downstream signaling. In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-protein-domain-split)
__ - __\@ref(fig:fig-fgfr4-position)__, __Table \@ref(tab:tab-fgfr-class)__). It can be hence speculated, that somatic mutations of the _FGFR_ genes may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling. In a more detailed analysis of mutation hot spots of the _FGFR3_ gene, mutation resulting in changes of protein residues S249C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'S249C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`% of cancer samples), Y373C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'Y373C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%), and R248C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'R248C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%) were present in $\geq$ 1% of cancer samples in at least two out of the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts (__Figure \@ref(fig:fig-fgfr3-hotspots)__).

@@

* At mRNA level, _FGFR3_, _TGFBR3_, and _FGFR1_ were the strongest expressed receptor genes. _FGF7_, _FGF11_, followed by _FGF2_ were the most abundant ligand transcripts; _HSPG2_, _SDC1_, and _CD44_ were the binding protein genes with the highest mRNA counts in the urothelial cancer tissue. Out of `r globals[c("receptors", "ligands", "binding_proteins")] %>% reduce(union) %>% length` genes of interest, `r ncol(hpa$ihc_score) - 1` proteins were quantified by immunohistochemistry (IHC) in the Human Protein Atlas (HPA). The protein quantities were expressed as IHC scores defined as products of staining intensity (1: negative to 4: strong) and quantity of positive cells (1: none to 4: >75%). In the HPA data set, `r `r sum(colMedians(hpa$ihc_score[-1]) > 1)` FGFR, FGF, and FGFBP proteins had median IHC scores $\geq$ 2, and were hence detectable in cancer specimens. At the protein/IHC level, FGFR1, FGFR3, FGFRL1, and FGFR4 were the most abundant receptors. FGF7, FGF10, FGF3, FGF19, and FGF17 were the strongest expressed ligand proteins. SDC2, SDC1, GPC4, CD44, and SDC4 were the most abundant FGFBP proteins (__Figures \@ref(fig:fig-expression)__ - __\@ref(fig:fig-ihc-fgfbp)__, __Table \@ref(tab:tab-expression)__).

* Co-expression of FGFR-, FGF-, and FGFBP-coding genes was explored by correlation and network analyses, which revealed a high degree co-regulation of the ligands, receptors, and binding proteins in the urothelial cancer tissue (__Figure \@ref(fig:fig-fgfr-correlation)__ and __Figure \@ref(fig:fig-fgfr-networks)__). Especially the network analyses point towards strong co-regulation of expression of _FGFR1_ with genes of its ligands (_FGF2_, _FGF7_, _FGF10_), and co-receptors such as _SDC2_, _DCN_ or _HSPG2_. Importantly, _FGFR1_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, general correlation strength with maximum of 1, degree: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), _FGF2_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% min`), _FGF7_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), and _FGF10_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) were identified as candidate highly connected hubs of the co-expression networks, which suggests that FGFR1 signaling as a critical player in urothelial cancer biology. In turn, _FGFR2_ was found to be weakly co-regulated with the non-canonical FGF receptor gene, _TGFBR3_. _FGFR3_ levels correlated weakly with levels of _SDC1_ and _GPC1_ transcripts that code for proteoglycan FGF binding proteins.

@@

* Analysis of differential expression of FGFR-, FGF-, and FGFBP-coding genes was conducted for the major consensus molecular subtypes of the TCGA BLCA, IMvigor, and BCAN cohort. Transcript levels of n = `r length(unique(unlist(sub_dge$common_significant)))` out of `r length(reduce(sub_dge$variables, union))` investigated genes (__Table \@ref(tab:tab-cons-dge)__). Among them, _FGFR3_, _SDC1_, _TGFBR3_, _GPC3_, and _SDC3_ were upregulated specifically in luminal papillary cancers. Overexpression of _KL_, a gene coding for [Klotho](https://www.genecards.org/cgi-bin/carddisp.pl?gene=KL) FGF-binding protein, was characteristic for luminal unstable tumors. Genes coding for FGFR protein, its ligands (_FGF7_, _FGF14_, _FGF10_, _FGF2_, _FGF9_, _FGF18_), and interaction partners (_DCN_, _HSPG2_, _SDC2_, _PTX3_) peaked in the stroma-rich class. High expression of _FGF5_, _FGF11_, and genes coding for binding proteins and FGF co-receptors, _CD44_, _FGFBP1_, _GPC1_ and _TNFAIP6_, was a hallmark of basal/squamous-like cancers (__Figures \@ref(fig:fig-sub-fgfr-expression)__ - __Figure \@ref(fig:fig-sub-fgf-expression2)__).

@@

* In summary, our analysis results suggest several modes of triggering of the oncogenic FGFR signaling highly specific for consensus molecular subsets of urothelial cancers (__Figure \@ref(fig:fig-summary)__). In papillary neoplasms, mutations of the _FGFR3_ gene affecting the hinge regions of the protein product and overexpression of co-receptors and binding proteins TGFB3, SDC1, GPC3, and SDC4 may culminate in a largely ligand-independent signaling. Such ligand-independent FGFR3 signaling was indeed reported for germ line mutations in the hinge region [@Moura2024]. In stroma-rich cancers, the tightly coordinated overexpression of FGFR1, its ligands (FGF2, FGF7, FGF9, FGF10, FGF14, FGF18) and interacting proteins (DCN, SDC1, HSPG2, ANOS1, and PTX3) may fuel ligand-dependent para- and autocrine signaling. In basal/squamous tumors, overexpression of FGF5 along with co-activators like FGFBP1, CD44, or GPC1, may trigger ligand-dependent signaling via FGFR1. Interestingly, the elevated FGFPB1 in basal/squamous cancers may overcome the low availability of other FGFR1 ligands, such as FGF2 or FGF5, by forming complexes with high affinity for the receptor [@Tassi2001; @Beer2005]. In luminal unstable cancers, overexpression of the _KL_ gene was observed, which codes for a co-receptor of FGF19, FGF21, and FGF23 [@Urakawa2006; @Goetz2007]. Yet, those FGF-coding genes were expressed in bulk cancer samples at nearly undetectable levels (TCGA BLCA, _FGF23_: median: `r median(2^(tcga$expression$FGF23) - 1)` counts, IQR: `r  quantile(2^(tcga$expression$FGF23) - 1, c(0.25, 0.75)) %>% paste(collapse = ' to ')` counts, 95^th^ percentile: `r quantile(2^(tcga$expression$FGF23) - 1, 0.95)` copies). Hence, the overexpression of _KL_ in the luminal unstable subset has likely no relevance for FGFR signaling. __Importantly, we could demonstrate that variability of expression of the FGFR-, FGF-, and FGFBP-coding genes emulates nearly completely the consensus molecular classification system by Kamoun et al. This reflects the immense importance of FGFR signaling for phenotypical variability of urothelial cancers. It also puts forward this signaling pathway as an Achilles' heel of all urothelial cancers independent of the molecular phenotype that can be exploited by FGFR inhibitors.__

@@

_(A) Transcript levels of FGFR-, FGF-, and FGFBP-coding genes were determined in cancer tissue samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` by RNA sequencing and published as $log_2$ gene counts (`r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`) or Z-scores of $log_2$ gene counts (`r globals$cohort_labs["bcan"]`). Median Z-scores of $log_2$ mRNA expression are shown in a heat map. Each tile represents median expression Z-score of a gene in one cohort. Numbers of cancer samples are indicated in the X axis. White tiles represent genes not detected in a given cohort._

@@

```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing. The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN). Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC). The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells. Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed. The table is available as a supplementary Excel file."}

@@

_Pairwise correlation of log2-transformed mRNA levels of FGF-, FGFR-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts by false discovery rate-adjusted Pearson's test. Correlation coefficients r for significant gene pairs are depicted in bubble plots. Color of the points codes for correlation sign. Point size represents absolute values of r. Numbers of cancer samples are displayed in the plot captions._

@@

```{r fig-fgfr-networks, fig.width = figur::convert(rep_figs$fgfr_networks, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.'}

@@

__Figure \@ref(fig:fig-fgfr-networks). Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.__

>>>>

* Five genetic and transcriptional collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (n = `r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (n = `r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers (n = `r nrow(bcan$clinic)`) [@Damrauer2022]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BCAN cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA, IMvigor, and BCAN cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, IMvigor, and BCAN cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Validation of expression of the genes of interest was done with immunohistochemistry data retrieved from the [Human Protein Atlas](https://www.proteinatlas.org/). Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__. In addition, analyses of expression, knockouts and RNA inference of FGFR-, FGF- and FGFBP-coding genes, as well as analyses resistance to pan-FGFR inhibitors were performed in n = `r nrow(depmap$cell_lines)`  urothelial cancer cell lines, whose data were retrieved from the [DepMap portal](https://depmap.org/portal).

@@

* Mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ were predominantly classified as missense mutations caused by single nucleotide polymorphisms (__Figure \@ref(fig:fig-mutation-type)__ and __\@ref(fig:fig-variant-type)__, __Table \@ref(tab:tab-fgfr-class)__).  For _FGFR1_, the mutation hot-spots were located in the Ig-like domain 1 and kinase domain. The kinase domain was the prime target of mutations of _FGFR2_. For _FGFR3_, they affected protein residues were located between Ig-like domains 1 and 2 responsible for ligand biding, or at the membrane-spanning hinge between the ligand-binding region and the kinase domain responsible for activation of the downstream signaling. In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-protein-domain-split)
__ - __\@ref(fig:fig-fgfr4-position)__, __Table \@ref(tab:tab-fgfr-class)__). In a more detailed analysis of mutation hot spots of the _FGFR3_ gene, mutation resulting in changes of protein residues S249C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'S249C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`% of cancer samples), Y373C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'Y373C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%), and R248C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'R248C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%) were present in $\geq$ 1% of cancer samples in at least two out of the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts (__Figure \@ref(fig:fig-fgfr3-hotspots)__). It can be hence speculated, that somatic mutations of the _FGFR_ genes may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling. In particular, constitutive, ligand-independent signaling was described for the most frequent R248C [@Naski1996], S249C [@Tomlinson2007], Y373C [@Bonaventure2007], G370C, and S371C [@Adar2002] variants of the FGFR3 protein.

@@

* At mRNA level, _FGFR3_, _TGFBR3_, _FGFR1_, and _FGFR2_ were the strongest expressed receptor genes. _FGF7_, _FGF11_, followed by _FGF2_ were the most abundant ligand transcripts; _HSPG2_, _SDC1_, and _CD44_ were the binding protein genes with the highest mRNA counts in the urothelial cancer tissue. _FGFR1_, _FGFR2_ and _FGFR3_ receptor genes, as well as _SDC1_, _HSPG2_, _CD44_, _SDC4_, _GPC1_ and _FIBP_ binding protein genes were also abundantly expressed at mRNA level by urothelial cancer cell lines in the DepMap repository [@DepMap2024] (__Figure \@ref(fig:fig-expression)A__, __Table \@ref(tab:tab-expression)__). This later observation shows clearly, that malignant epithelium cells are equipped with the key molecules of the oncogenic FGFR signaling. Out of `r globals[c("receptors", "ligands", "binding_proteins")] %>% reduce(union) %>% length` genes of interest, `r ncol(hpa$ihc_score) - 1` proteins were quantified by immunohistochemistry (IHC) in the Human Protein Atlas (HPA). The protein quantities were expressed as IHC scores defined as products of staining intensity (1: negative to 4: strong) and quantity of positive cells (1: none to 4: >75%). In the HPA data set, `r `r sum(colMedians(hpa$ihc_score[-1]) > 1)` FGFR, FGF, and FGFBP proteins had median IHC scores $\geq$ 2, and were hence detectable in cancer specimens. At the protein/IHC level, FGFR1, FGFR3, FGFRL1, and FGFR4 were the most abundant receptors. FGF7, FGF10, FGF3, FGF19, and FGF17 were the strongest expressed ligand proteins. SDC2, SDC1, GPC4, CD44, and SDC4 were the most abundant FGFBP proteins (__Figures \@ref(fig:fig-expression)B__ - __\@ref(fig:fig-ihc-fgfbp)__, __Table \@ref(tab:tab-expression)__).

* Co-expression of FGFR-, FGF-, and FGFBP-coding genes in the urothelial cancer tissue was explored by correlation and network analyses, which revealed a high degree co-regulation of the ligands, receptors, and binding proteins in the urothelial cancer tissue (__Figure \@ref(fig:fig-fgfr-correlation)__ and __Figure \@ref(fig:fig-fgfr-networks)__). Especially the network analyses point towards strong co-regulation of expression of _FGFR1_ with genes of its ligands (_FGF2_, _FGF7_, _FGF10_), and co-receptors such as _SDC2_, _DCN_ or _HSPG2_. Importantly, _FGFR1_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, general correlation strength with maximum of 1, degree: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), _FGF2_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% min`), _FGF7_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), and _FGF10_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) were identified as candidate highly connected hubs of the co-expression networks, which suggests that FGFR1 signaling as a critical player in urothelial cancer biology. In turn, _FGFR2_ was found to be weakly co-regulated with the non-canonical FGF receptor gene, _TGFBR3_. _FGFR3_ levels correlated weakly with levels of _SDC1_ and _GPC1_ transcripts that code for proteoglycan FGF binding proteins.

* A similar co-expression network analysis was conducted for FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer cell lines in the DepMap repository. Its results revealed two separate communities with, respectively, _FGFR1_ and _FGFR2/3_ as candidate hubs. As described above for the urothelial cancer specimens, the gene community of _FGFR1_ included also genes of FGFR1 ligands (_FGF1_, _FGF2_, _FGF5_, _FGF7_, _FGF9_) and co-receptors (_CD44_, _FGFRL1_, _TNFAIP6_) (__Figure \@ref(fig:fig-cell-networks)__). This suggests that urothelial cancer cells employ either a ligand-dependent signaling via FGFR1 or a possibly ligand-independent signaling via FGFR2/3. Those signaling pathways are likely non-redundant.

* Next, we sought to investigate if urothelial malignant cells depend on FGFR signaling by an analysis of CRISPR/Cas9 and RNA interference screening data for urothelial cancer cell lines in the DepMap repository [@DepMap2024; @McFarland2018; @Dempster2021]. In this analysis, we compared effects of knockouts and RNA silencing of genes coding for FGF receptor proteins with background effects for genes without meaningful expression in the urothelial cancer cell lines. Growth of JMSU-1, UM-UC-13, UM-UC-11, TCCSUP, and 639-V cell lines was significantly inhibited by knockout or silencing of the _FGFR1_ gene, which points towards critical dependence on FGFR1 signaling. Analogically, the UM-UC-1 cell line growth was inhibited by the _FGFR2_ CRISPR/Cas9 knockout. UM-UC-14, RT-112, RT112/84, and SW 780 cell line proliferation was curbed by the knockout or silencing of the _FGFR3_ gene. Notably, the effects of _FGFR1_, _FGFR2_, and _FGFR3_ gene knockouts or interference were independent of _FGFR3_ mutation status of a cell line (__Figure \@ref(fig:fig-crispri)__).

* In a similar analysis, we delved into resistance of DepMap urothelial cancer cell lines to pan-FGFR inhibitors in the GDSC1/2 (compounds: `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`) [@Yang2013] and PRISM drug screening experiments (compounds: `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`) [@Yu2016]. The resistance was metered in the GDSC1/2 experiments by IC50, a concentration of a compound that results in 50% cell line growth inhibition; sensitive cell lines were identified by IC50 < 10 µM, i.e. the highest concentration of the FGFR inhibitors In the PRISM experiment, the resistance was gauged by AUC, an area under the compound's dose-response curve normalized for response of the DMSO-treated cells; sensitive cell lines were identified by AUC < 1. As presented in __Figure \@ref(fig:fig-resistance)__, urothelial cancer cell lines responded to pan-FGFR inhibitors in both the GDSC1/2 and PRISM experiments in a _FGFR3_ mutation-independent manner. _FGFR3_ wild-type cell lines RT-112, RT4, SW 780, CAL-29, and J82 showed particularly high sensitivity to multiple FGFR blockers in the GDSC1/2 experiments. Erdafitinib is the sole clinically approved pan-FGFR inhibitor in urothelial cancer. _FGFR3_ wild-type cell lines `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$name_lab %>% collapse_and` demonstrated an especially good sensitivity to Erdafitinib in the PRISM assay (AUC: `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`). By contrast, `r expl_res$sensitive$prism %>% filter(.data[['PRISM_BRD-K84868168-001-01-0_erdafitinib']] == 'resistant') %>% .$sample_id %>% exchange(depmap$cell_lines, key = 'sample_id', value = 'cell_line_name') %>% collapse_and` cell lines were found to be resistant to erdafitinib with AUC values ranging from `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), status == 'resistant') %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`. Collectively, the pan-FGFR inhibitor response data, and the CRISPR/Cas9 and RNA interference screening results indicate FGFR signaling addiction of urothelial cancer cells independently of _FGFR3_ mutations. They also argue for a broader use of pan-FGFR inhibitors in urothelial cancer patients without _FGFR3_ gene alterations.

@@

* Analysis of differential expression of FGFR-, FGF-, and FGFBP-coding genes was conducted for the major consensus molecular subtypes of the TCGA BLCA, IMvigor, and BCAN cohort. Transcript levels of n = `r length(unique(unlist(sub_dge$common_significant)))` out of `r length(reduce(sub_dge$variables, union))` investigated genes were significantly differentially regulated between consensus molecular subtyped in at least two cohorts (__Table \@ref(tab:tab-cons-dge)__). Among them, _FGFR3_, _SDC1_, _TGFBR3_, _GPC3_, and _SDC3_ were upregulated specifically in luminal papillary cancers. Overexpression of _KL_, a gene coding for [Klotho](https://www.genecards.org/cgi-bin/carddisp.pl?gene=KL) FGF-binding protein, was characteristic for luminal unstable tumors. Genes coding for FGFR protein, its ligands (_FGF7_, _FGF14_, _FGF10_, _FGF2_, _FGF9_, _FGF18_), and interaction partners (_DCN_, _HSPG2_, _SDC2_, _PTX3_) peaked in the stroma-rich class [@Ferguson2021]. High expression of _FGF5_, _FGF11_, and genes coding for binding proteins and FGF co-receptors, _CD44_, _FGFBP1_, _GPC1_ and _TNFAIP6_, was a hallmark of basal/squamous-like cancers (__Figures \@ref(fig:fig-sub-fgfr-expression)__ - __Figure \@ref(fig:fig-sub-fgf-expression2)__).

@@

* In summary, our analysis results underline the biological relevance of FGFR signaling as an oncogenic driver of urothelial carcinoma irrespectively of _FGFR3_ gene alterations. FGFR1 and FGFR3 emerge as the main hubs of this signaling pathway. Additionally, we propose several modes of triggering of the oncogenic FGFR signaling highly specific for consensus molecular subsets of urothelial cancers (__Figure \@ref(fig:fig-summary)__). In papillary neoplasms, mutations of the _FGFR3_ gene affecting the hinge regions of the protein product [@Naski1996; @Tomlinson2007; @Bonaventure2007; @Adar2002; @Moura2024] and, possibly, overexpression of co-receptors and binding proteins TGFB3, SDC1, GPC3, and SDC4 [@Huttlin2021] may culminate in a largely ligand-independent signaling. In stroma-rich cancers, the tightly coordinated overexpression of FGFR1, its ligands (FGF2, FGF7, FGF9, FGF10, FGF14, FGF18) and interacting proteins (DCN, SDC1, HSPG2, ANOS1, and PTX3) may fuel ligand-dependent para- and autocrine signaling [@Ferguson2021]. In basal/squamous tumors, overexpression of FGF5 along with co-activators like FGFBP1, CD44, or GPC1, may trigger ligand-dependent signaling via FGFR1 [@Ferguson2021]. Interestingly, the elevated FGFPB1 in basal/squamous cancers may overcome the low availability of other FGFR1 ligands, such as FGF2 or FGF5, by forming complexes with high affinity for the receptor [@Tassi2001; @Beer2005]. In luminal unstable cancers, overexpression of the _KL_ gene was observed, which codes for a co-receptor of FGF19, FGF21, and FGF23 [@Urakawa2006; @Goetz2007]. Yet, those FGF-coding genes were expressed in bulk cancer samples at nearly undetectable levels (TCGA BLCA, _FGF23_: median: `r signif(median(2^(tcga$expression$FGF23) - 1), 2)` counts, IQR: `r  quantile(2^(tcga$expression$FGF23) - 1, c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' to ')` counts, 95^th^ percentile: `r signif(quantile(2^(tcga$expression$FGF23) - 1, 0.95), 2)` copies). Hence, the overexpression of _KL_ in the luminal unstable subset has likely no relevance for FGFR signaling. __Importantly, we could demonstrate that variability of expression of the FGFR-, FGF-, and FGFBP-coding genes emulates nearly completely the consensus molecular classification system by Kamoun et al. This reflects the immense importance of FGFR signaling for phenotypical variability of urothelial cancers. It also puts forward this signaling pathway as an Achilles' heel of all urothelial cancers independent of the molecular phenotype that can be exploited by pan-FGFR inhibitors.__

@@

_(A) Transcript levels of FGFR-, FGF-, and FGFBP-coding genes were determined in cancer tissue samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` by RNA sequencing and published as $log_2$ gene counts (`r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`) or Z-scores of $log_2$ gene counts (`r globals$cohort_labs["bcan"]`). mRNA expression of the genes of interest in DepMap urothelial cancer cell lines was measured by RNA sequencing and provided as $log_2$ gene counts. Median Z-scores of $log_2$ mRNA expression are shown in a heat map. Each tile represents median expression Z-score of a gene in one cohort. Numbers of cancer samples are indicated in the X axis. White tiles represent genes not detected in a given cohort._

@@

```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohorts by RNA sequencing. The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN). Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC). The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells. Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed. The table is available as a supplementary Excel file."}

@@

_Pairwise correlation of log2-transformed mRNA levels of FGF-, FGFR-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts og bulk cancer transcriptomes, and in DepMap urothelial cancer cell lines. Statistical significance was determined by false discovery rate (FDR) adjusted Spearman's permutation test. Significant effects were considered for pFDR < 0.05 and $abs(\rho)\geq$ 0.3. Correlation coefficients $\rho$ for gene pairs found to be significant in all four collectives are depicted in bubble plots. Color of the points codes for correlation sign. Point size represents absolute value of $\rho$. Numbers of cancer samples and cell lines are displayed in the plot captions._

@@

```{r fig-fgfr-networks, fig.width = figur::convert(rep_figs$fgfr_networks, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes: cancer tissue.'}

@@

__Figure \@ref(fig:fig-fgfr-networks). Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes: cancer tissue.__

_Co-expression networks were constructed in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` bulk cancer transcriptome cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman's rho $\rho \geq$ 0.3. Isolated vertices of the networks were removed. The networks were visualized using the Fruchterman-Reingold algorithm. Points represent node vertices/genes, point shape codes for classification of the protein product. Point color represent hub scores, which measure the overall correlation strength. Lines represent edges, i.e. correlations between expression levels with $\rho \geq$ 0.3. Line widths correspond to $\rho$ values. Note the central position, connectivity, and high hub scores of the genes coding for FGFR1, its ligands FGF2, FGF7 and FGF10, and the binding proteins SDC2, DCN, PTX3 and HSPG2 preserved in all cohorts. This suggests that signaling via FGFR1 may serve as a major driver of urothelial cancer._
